---
title: 'dHPC gene expression 6 hours post-LPS: Array QC'
author: "G. Lewandowski"
date: "February 10, 2017"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Generation of a GeneFeatureSet container for the **6 hours** LPS-Microarry data and sample information.

### GeneFeatureSet container generation using the `oligo` package.  
Set up GeneFeatureSet container for the LPS_Microarray data.  Read in sample information and use the `oligo` package to read in the raw intensity data (CEL files). 

### Load CEL files using the `oligo` package and add the sample info as phenoData to the `GeneFeatureSet` object.

```{r GeneFeatueSet lpsrawdata, eval=FALSE}
# load the oligo package
suppressPackageStartupMessages(library(oligo))

# list the CEL files
# the CEL file names will be used as rownames in the phenoData and as sampleNames
celfiles <- list.celfiles(file.path("LPS_Data_Files"))

# set up sample info data frame by 
# (1) reading in the sample_info.txt file (returns a data.frame object)
# (2) change the rownames to the CEL filenames (sample names),
# (3) remove the first (redundant) column

sampleInfo <- read.delim(file = file.path("LPS_Data_Files", "sample_info.txt"), 
                         header = TRUE, stringsAsFactors = FALSE)
# order the sampleInfo according to CEL file name (column x)
sampleInfo <- sampleInfo[order(sampleInfo$X),]
# change rownames from 1-12 to sample names
rownames(sampleInfo) <- substr(celfiles, 1, 5)
# change column 1 name from X to filename
colnames(sampleInfo)[1] <- "filename" 

# set up metadata with row names matching the column names of the sampleInfo df
# because these are single channel arrays, use ALL for channel value
rns <- colnames(sampleInfo)
metadata <- data.frame(labelDescription = rns, channel = factor('ALL'))

# load the CEL files; returns a "GeneFeatureSet" object
lpsrawdata <- read.celfiles(list.celfiles(file.path("LPS_Data_Files"), 
                                          full.name = TRUE)) 

# add the sample info as phenoData to lpsdataraw
# set up the pheno data for lpsdataraw
pd <- new('AnnotatedDataFrame', data = sampleInfo, varMetadata = metadata)

# set the phenoData for lpsdataraw
phenoData(lpsrawdata) <- pd

# set the sample names in the lpsrawdata container to the sample names from CEL file names
sampleNames(lpsrawdata) <- rownames(sampleInfo)

# Subset the data to obtain the 6 hour treatment samples:
lpsrawdata.6h <- lpsrawdata[, lpsrawdata$group %in% c("sal.6h", "lps.6h")]
rm(pd, rns, sampleInfo, metadata)
```

### Save the lpsrawdata GeneFeatureSet object to an external file.

```{r save GeneFeatureSet object, eval=FALSE, results='hide'}
save(lpsrawdata, file = file.path("processedData", "lpsrawdata.Rdata"))
save(lpsrawdata.6h, file = file.path("processedData", "lpsrawdata_6h.Rdata"))
```

```{r remove lpsrawdata, eval=FALSE, echo=FALSE}
rm(lpsrawdata)
```


```{r load lpsrawdata.6h, echo=FALSE}
# load the oligo package
suppressPackageStartupMessages(library(oligo))

load(file.path("processedData", "lpsrawdata_6h.Rdata"))
```


## Quality control analysis of the experimental arrays

### Intensity distribution across the arrays:

#### Intensity (density) Histograms

```{r intensity histograms, cache=TRUE, fig.align='center', fig.height=4, fig.width=6}

library(RColorBrewer)
par(mfrow = c(1,2))
usr.col <- brewer.pal(12, "Paired")
set.seed(1234)
oligo::hist(lpsrawdata.6h, which = "all", lty = rep(1, 6), 
            col = usr.col, main = "Intensity Histograms")

legend("topright", sampleNames(lpsrawdata.6h), lty = rep(1, 6), 
       col = usr.col, cex = 0.6)
```

The intensity distributions for all arrays are similiar.    

#### Box plots:

```{r intensity distribution boxplots, cache=TRUE, fig.align='center', fig.height=4, fig.width=6}

set.seed(1234)
oligo::boxplot(lpsrawdata.6h, which = "all", 
               col = usr.col, las =3, cex.axis = 0.75, 
               names = sampleNames(lpsrawdata.6h), main = "Intensity Distribution")
```

The median intensity and widths of the intensity distributions are similar for all 6 arrays.  

### MA plots
Plotting log-ratios, **M**, versus average log-intensities, **A** for two arrays (samples) is a strategy to visualize the relationship between two sample. For the single channel Affy Gene ST arrays, the comparison is done with pairs of samples.  If two samples are from the same experimental group, then the arrays should behave in a similar manner.  Meaning that the plotted data points should appear symmetrically about a horizontal line through zero.  Any deviation from the zero horizontal line represents different responses of the two samples.  MA plots of the raw (non-normalized) data can be used to determine if there are any array problems.  MA plots are done for pairs in each of the four groups. The groups are: "sal.6h" "lps.6h".    

Normalization  efforts are then used to get the data points on the MA plot to be fitted symmetrically around the zero horizontal line.  

```{r MA plots, fig.align='center'}

# relevel groups so that sal.6h is the reference group
lpsrawdata.6h$group <- factor(lpsrawdata.6h$group)
lpsrawdata.6h$group <- relevel(lpsrawdata.6h$group, ref = "sal.6h")
groups <- (lpsrawdata.6h$group)

xl = c(2.8, 4)
yl = c(-1, 1)

# MA plot for the saline control group at 6 hours (rats 1,3,5)
MAplot(lpsrawdata.6h[, groups == "sal.6h"], main = "saline",
              pairs=TRUE, ylim = yl, xlim = xl)

# MA plot for the lps treated group at 6 hours (rats 2,4,24)
MAplot(lpsrawdata.6h[, groups == "lps.6h"], pairs=TRUE, 
       ylim = yl, xlim = xl, main = "LPS")

# MA plot for 6 h lps vs reference 6 h saline
MAplot(lpsrawdata.6h, groups=groups, which=2, 
       refSamples=1, show.statistics=FALSE)
```

### Probe level models
Looking for probe effects in the raw data; PLM fits a model and outputs a number of statistics that are used for quality diagnostics.

```{r PLMs, cache=TRUE}
# Raw data no preprocessing
# summarization target = 'core', background correction = F, normalization = F

# load the plm for lpsrawdata.6h if it already exists, else, generate the plm for the raw data
plm_file <- file.path("processedData", "plm_lpsrawdata_6h.RData")

if (file.exists(plm_file)){
    load(plm_file)
} else {
    plm_fit <- fitProbeLevelModel(lpsrawdata.6h, target = "core", 
                                  background = FALSE, normalize = FALSE)
    save(plm_fit, file = plm_file)
}

```

```{r remove, echo=FALSE}
rm(plm_file, groups)
```


#### Relative log expression and Normalized Unscaled Standard Error:
**Relative Log Expression (RLE)** plot is constructed from the PLM model.  For each array the RLE plot is a box plot of the relative expression for each gene- which is defined as the difference of the PLM estimated gene expression an the median expression.  

**Normalized Unscaled Standard Error (NUSE)** for each gene is defined as the PLM fit standard error minus the median standard error across all the arrays.  The median standard error for any probeset should be 1 across all the arrays.  A NUSE value significantly higher from other samples likely indicates a lower quality chip (array).

```{r RLE Nuse boxplots, cache=TRUE}
par(mfrow = c(1,2))

# box plots of RLE values from the raw data
RLE(plm_fit, type = "plot",las =3, cex.axis = 0.75, 
    names = sampleNames(lpsrawdata.6h), main = "RLE (raw data)")

# box plots of NUSE values from the raw data
NUSE(plm_fit, type = "plot",las =3, cex.axis = 0.75, 
     names = sampleNames(lpsrawdata.6h), main = "NUSE (raw data)")
```

#### R session information:
```{r session info, echo=FALSE}
sessionInfo()
suppressMessages(gc())
```
